Query Optimization for Dynamic Imputation

### 2017-03-01 paper/algorithm.tex

Each imputation operator incurs a heuristic loss penalty. We define \textsc{Loss} as the average over the individual penalties incurred
by $\mu,\delta$ operators. Our recursive definition of $L$ accumulates these penalties, and $N$ counts the number of imputation operations.
For example, $\delta$ contributes a loss of 1 (i.e. we lose all information  that dropped tuples
might have encoded). Meanwhile, imputing \nullv{} fields with $\mu$ incurs a loss penalty $p \in (0, 1]$. In this case, the penalty is
inversely proportional to the square root of the number of cells available to train the imputation algorithm, which is defined as
as the product of the number of attributes and the number of tuples. Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information lost \textit{decreases} with more values to work
with. We wanted to encode the fact that the number of training observations has diminishing returns for imputation quality,
and found that scaling the cell count by applying square root worked well in practice.

@@

Scan costs are estimated based on the number of tuples, 
page size, and an unit-less I/O cost per page. Join costs are estimated as a function
of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
Filtering, projecting, and aggregating are all done in memory and have low computational overhead
so we assume negligible time costs for those operations.

@@

In general, evaluating the time complexity for different \textit{Impute} operators is
complicated, as the properties of
different algorithms can vary significantly, the underlying I/O cost
is not easily extracted (especially in the case that the entire set of tuples does not
fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
not even consider CPU cost explicitly). For example, \cite{martin1995time} find that the
time complexity of the \textit{build tree} algorithm for one commonly-used class of
decision trees is a function of the number of classes, several overhead constants, the
parameterization of the partition and heuristic functions, and the \textit{arity} (the
number of subsets considered for each split). The \textit{build tree} phase often
does not even dominate computation, as post-processing steps like pruning, which are vital
in achieving good performance, can have cost exponential in the height of the tree. In
practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
plans).

>>>>

Each imputation operator incurs a heuristic loss penalty. We define \textsc{Loss} as the average over the individual penalties incurred
by $\mu,\delta$ operators. Our recursive definition of $L$ accumulates these penalties, and $N$ counts the number of imputation operations.
For example, $\delta$ contributes a loss of 1 (i.e. we lose all information  that dropped tuples
might have encoded). Meanwhile, imputing \nullv{} fields with $\mu$ incurs a loss penalty $p \in (0, 1]$. In this case, the penalty is
inversely proportional to the square root of the number of cells available to train the imputation algorithm, which is defined as
the product of the number of attributes and the number of tuples. Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information lost \textit{decreases} with more values to work
with. We wanted to encode the fact that the number of training observations has diminishing returns for imputation quality,
and found that scaling the cell count by applying square root worked well in practice.

@@

Scan costs are estimated based on the number of tuples, 
page size, and a unit-less I/O cost per page. Join costs are estimated as a function
of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
Filtering, projecting, and aggregating are all done in memory and have low computational overhead
so we assume negligible time costs for those operations.

@@

In general, evaluating the time complexity for different \textit{Impute} operators is
complicated, as the properties of
different algorithms can vary significantly, the underlying I/O cost
is not easily extracted (especially in the case that the entire set of tuples does not
fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
not even consider CPU cost explicitly). For example,~\cite{martin1995time} find that the
time complexity of the \textit{build tree} algorithm for one commonly-used class of
decision trees is a function of the number of classes, several overhead constants, the
parameterization of the partition and heuristic functions, and the \textit{arity} (the
number of subsets considered for each split). The \textit{build tree} phase often
does not even dominate computation, as post-processing steps like pruning, which are vital
in achieving good performance, can have cost exponential in the height of the tree. In
practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
plans).

### 2017-03-01 paper/conclusion.tex

\section{Conclusion}
Our evaluation of \ProjectName{} shows that missing values and their imputation can be successfully integrated into the relational calculus and
existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
learning technique, as operators in the algebra and use a simple, but effective, cost model to consider tradeoffs in 
imputation quality and runtime. We showed that different values for the tradeoff parameter can yield substantially
different plans, allowing the user to express their preferences for performance on a per-query basis.
Our experiments with \textit{CDC}, \textit{freeCodeCamp} and \textit{ACS} survey-based datasets
show that \ProjectName{} can be used successfully with real-world data, expressing standard SQL queries over data with missing values. 
We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
the data exploration phase. The plans selected for each query execute an order-of-magnitude faster than
 the standard approach of imputing on the entire base tables and
then formulating queries. Furthermore, the difference in query results between the two approaches is
shown to be small in all queries considered\todo{quantify this}. Data analysts need not commit to a specific imputation strategy and can instead
vary this across queries.

>>>>

\section{Conclusion}
Our evaluation of \ProjectName{} shows that missing values and their imputation can be successfully integrated into the relational calculus and
existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
learning technique, as operators in the algebra and use a simple, but effective, cost model to consider trade-offs in 
imputation quality and runtime. We showed that different values for the trade-off parameter can yield substantially
different plans, allowing the user to express their preferences for performance on a per-query basis.
Our experiments with \textit{CDC}, \textit{freeCodeCamp} and \textit{ACS} survey-based datasets
show that \ProjectName{} can be used successfully with real-world data, expressing standard SQL queries over data with missing values. 
We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
the data exploration phase. The plans selected for each query execute an order-of-magnitude faster than
 the standard approach of imputing on the entire base tables and
then formulating queries. Furthermore, the difference in query results between the two approaches is
shown to be small in all queries considered\todo{quantify this}. Data analysts need not commit to a specific imputation strategy and can instead
vary this across queries.
